Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands | Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins

### 2024-12-16 MANUSCRIPT.qmd
draft

abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base flow processes and provide a framework for water resource management in data-limited regions.

@@

Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge [@usgs-glossary-2018]. Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals. Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020]. Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013]. Effective management of water quantity and quality depends on understanding seasonal and interannual base flow patterns and long-term changes in base flow behavior.

@@

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in baseflow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded. These gages were identified through annual reports on water data published by the USGS [@usgs2010]. Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

@@

Statistical analyses were conducted on annual BFI and base flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations. This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].

@@

Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining baseflows. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to baseflow declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to baseflow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|------------|------------|------------|------------|------------|------------|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

@@

Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in baseflow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in baseflow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].

The analysis shows that baseflow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge. Coincidence with AET (actual evapotranspiration) trends (55.12% for baseflow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or baseflow during dry periods. In contrast, temperature trends show lower percentages of coincidence, often opposing baseflow and BFI trends. Specifically, positive (negative) temperature trends are frequently associated with negative (positive) baseflow trends (60.98%) and BFI trends (52.68%). These results highlight the complex interplay between climatic variables and hydrological processes. While precipitation exerts the strongest influence on baseflow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from [@beck2013], highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features. Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in \@fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.

@@

-   This regional model, trained on a relatively sparse dataset over a region spanning 295,253 km^2^, is shown to predict BFI well spatially and across climate classifications.

-

>>>>

abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.

@@

Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge [@usgs-glossary-2018]. Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals. Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020]. Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013]. Effective management of water quantity and quality depends on understanding seasonal and interannual base-flow patterns and long-term changes in base-flow behavior.

@@

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded. These streamgages were identified through annual reports on water data published by the USGS [@usgs2010]. Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

@@

Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations. This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].

@@

Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining base flow. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|----|----|----|----|----|----|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

@@

Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends). This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].

The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge. Coincidence with AET (actual evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods. In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends. Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%). These results highlight the complex interplay between climatic variables and hydrological processes. While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|----|----|----|----|----|----|----|----|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from [@beck2013], highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022]. This relationship is further supported by hydroclimate features. Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence. In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.

@@

This study provides new insights into base-flow dynamics and groundwater contributions in Arizona’s dryland rivers by combining an analysis of instrumented streamflow records with machine learning predictions for ungauged basins. The results highlight significant spatial variability in BFI, with approximately 32% of Arizona’s long-term streamflow originating from groundwater discharge. Regions such as the Grand Canyon and Mogollon Rim demonstrate high BFI values due to strong groundwater-surface water interactions, while areas like the Little Colorado River Basin exhibit low BFI values, reflecting limited groundwater recharge.

Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity. These findings emphasize the critical role of climate-hydrology interactions in shaping groundwater contributions to streamflow. Inverse trends between temperature and BFI suggest that further warming could reduce groundwater contributions to streamflow. Coincident trends in precipitation and BFI further underscore the importance understanding recharge processes in arid and semi-arid landscapes.

Using an XGBoost machine learning algorithm, we successfully predicted long-term BFI in ungauged basins, achieving strong model performance (R² = 0.764, RMSE = 0.129). The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics. Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models. These predictions address the limitations posed by Arizona’s sparse streamgage network, offering a scalable approach to estimate BFI in data-limited regions.

DATA LIMITATIONS - ABE

This study demonstrates the benefits of combining observational records with machine learning to improve our understanding of streamflow processes in drylands. Future work should explore the projected effects of climate change on base-flow processes and developing models for other data-poor regions. The framework presented here has broad applicability to other arid and semi-arid regions worldwide and can inform water resource management strategies aimed at addressing water scarcity and adapting to climate variability.
